---
author: "Luke Ring"
date: "`r format(Sys.time(), '%d %B, %Y')`"
title: "CDS: Practicing Functions with Gapminder"
output:
  html_document:
    css: gapminder.css
    highlight: breezedark
    theme: darkly
    toc: yes
    toc_depth: 2
    # number_sections: true
    #toc_float:
    #  collapsed: yes
    #  smooth_scroll: yes
    fig_height: 7
    fig_width: 10
    fig_retina: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(gapminder)
library(ggthemes)

```

This is the assignment for Cultural Data Science, Week 43. Assignment 6.

# Assignment

Use the gapminder dataset from Week 43 to produce solutions to the three tasks below. 

## Task 1: Defensive GDP Function

> Define a defensive function that calculates the Gross Domestic Product of a nation from the data  available in the gapminder dataset. You can use the population and GDPpercapita columns for it. Using that function, calculate the GDP of Denmark in the following years: 1967, 1977, 1987, 1997, 2007, and 2017.

First we define the function:

```{r task1}
# simple function to test if a value is interger-like
is_wholenumber <- function(x) {
    if (is.numeric(x) & x %% 1 == 0) {
        return(TRUE)
    } else {
        return(FALSE)
    }
}

# The main GDP calculation function
gapminder_gdp <- function(for_country, for_year) {
    # first check that country is a character
    if (!is.character(for_country)) {
        stop("country must be a character")
    }

    # then check that year is an integer
    if (!is_wholenumber(for_year)) {
        stop("year must be an integer")
    }

    # ensure dplyr is loaded
    if (!require(dplyr)) {
        stop(paste(
            "dplyr is required to run this function.",
            "Please install it with",
            "install.packages('dplyr')"
        ))
    }

    # make sure the gampnider dataset exists
    if (!require(gapminder)) {
        stop(paste(
            "Gapminder package not available,",
            "you can install by running",
            'install.packages("gapminder")'
        ))
    }

    # make the country and gapminder countries lowercase
    # to avoid case sensitivity issues
    country_orig <- for_country
    for_country <- tolower(for_country)
    levels(gapminder$country) <- tolower(levels(gapminder$country))

    # check that the country is in the gapminder dataset
    if (!for_country %in% gapminder$country) {
        stop(paste(
            "Country",
            country_orig,
            "not found in gapminder dataset,",
            "please check the spelling of the country"
        ))
    }

    # check that the year is in the gapminder dataset
    if (!for_year %in% gapminder[gapminder$country == for_country, ]$year) {
        # we can report available range of years for the country
        # to help the user
        years <- gapminder %>%
            filter(country == for_country) %>%
            pull(year)
        stop(paste(
            "Year",
            for_year,
            "not found in gapminder dataset for",
            paste0(country_orig, "."),
            "\nPlease ensure the year is between",
            min(years), "and", paste0(max(years), ".")
        ))
    }

    # calculate the GDP
    gdp <- gapminder %>%
        filter(country == for_country, year == for_year) %>%
        mutate(gdp = pop * gdpPercap) %>%
        pull(gdp)

    # return the GDP
    return(gdp)
}
```

Now we can test the function output for the years requested:

```{r task1_test}
# test the function
test_country <- "Denmark"
test_years <- c(1967, 1977, 1987, 1997, 2007, 2017)

# loop through the years and print the results
for (test_year in test_years) {
    # get the GDP for the country and year
    tryCatch({
        gdp <- gapminder_gdp(test_country, test_year)
        # print the results
        print(paste(
            "The GDP of",
            test_country,
            "in",
            test_year,
            "was",
            scales::label_dollar(prefix = "US$")(gdp)
        ))
    }, error = function(e) {
        # print the error message
        print(e)
    })
}
```

We can see that the output shows the GDP for all the years except for 2017, which is not in the dataset. In this case an error is thrown which stops execution. This means you can't accidentally get strange results from the function, or if you try to do something that isn't possible, it will not just silently continue, but will let you know immediately, allowing you to fix the issue.

In the test examples above, we have a `try / catch` block which allows us to continue our script / Rmd knitting even if there is an error, this is fine for testing but you would not want to do that with real data unless you are absolutely sure that you don't care about data that couldn't be found.

## Task 2: Life Expectancy for countries starting with "B"

> Write a script that loops over each country in the gapminder dataset, tests whether the country starts with a 'B' , and prints out whether the life expectancy is smaller than 50, between 50 and 70, or greater than 70. (Hint: remember the grepl function, and review the Control Flow tutorial)

```{r task2}
# loop through the countries
for (cntry in unique(gapminder$country)) {
    # test if the country starts with a B
    if (!grepl("^B", cntry)) {
        # if not, skip to the next country
        next
    }

    # get the life expectancy
    life_expectancy <- gapminder %>%
        filter(country == cntry) %>%
        pull(lifeExp)

    # calculate the mean life expectancy for the years found
    mean_life_expectancy <- mean(life_expectancy, na.rm = TRUE)

    # get the range of years
    years <- gapminder %>%
        filter(country == cntry) %>%
        pull(year)

    # print the results
    print(paste(
        "The mean life expectancy for",
        cntry,
        "from",
        min(years), "to", max(years),
        "was: "
    ))

    # test the life expectancy
    if (mean_life_expectancy < 50) {
        print(paste("less than 50 years"))
    } else if (mean_life_expectancy >= 50 & mean_life_expectancy <= 70) {
        print(paste("between 50 and 70 years"))
    } else {
        print(paste("greater than 70 years"))
    }
}
```

### Bonus: the same, but using `tidyverse`

```{r task2_bonus}
# find countries starting with B, and get the mean life expectancy
gapminder %>%
    filter(grepl("^B", country)) %>%
    group_by(country) %>%
    summarise(
        mean_life_expectancy = mean(lifeExp, na.rm = TRUE),
        years = paste0(min(year), "-", max(year))
    ) %>%
    mutate(
        life_expectancy = case_when(
            mean_life_expectancy < 50 ~ "less than 50 years",
            mean_life_expectancy >= 50 &
                mean_life_expectancy <= 70 ~ "between 50 and 70 years",
            TRUE ~ "greater than 70 years"
        )
    ) %>%
    select(country, years, life_expectancy) %>%
    print(n = Inf)
```

## Task 3: Plot `life expectancy/time` for countries starting with "M"

> Challenge/Optional: Write a script that loops over each country in the gapminder dataset, tests whether the country starts with a 'M' and graphs life expectancy against time (using plot() function) as a line graph if the mean life expectancy is under 50 years.

```{r task3}
# loop through the countries
for (cntry in unique(gapminder$country)) {
    # test if the country starts with a M
    if (!grepl("^M", cntry)) {
        # if not, skip to the next country
        next
    }

    # get the life expectancy
    life_expectancy <- gapminder %>%
        filter(country == cntry) %>%
        pull(lifeExp)

    # calculate the mean life expectancy for the years found
    mean_life_expectancy <- mean(life_expectancy, na.rm = TRUE)

    # test the life expectancy
    if (mean_life_expectancy < 50) {
        # get the range of years
        years <- gapminder %>%
            filter(country == cntry) %>%
            pull(year)
        # plot the life expectancy and make it look nice
        par(bg = "#121212")
        plot(
            years,
            life_expectancy,
            type = "l", # line plot
            col = "#f0afff", # line color
            lwd = 2, # line width
            cex = 1.2, # label size multiplier
            xlab = "",
            ylab = "",
            axes = FALSE,
            ann = FALSE,
            oma = c(2, 3, 5, 2) # outer margins, b, l, t, r
        )
        # add title
        mtext(
            side = 3,
            line = 2,
            at = min(years) - 4,
            adj = 0,
            cex = 1.5,
            col = "#ececec",
            paste("Life Expectancy for", cntry),
            font = 2 # bold the title
        )
        # add subtitle
        mtext(side = 3,
            line = 0.75,
            at = min(years) - 4,
            adj = 0,
            cex = 1,
            col = "#ececec",
            paste("From", min(years), "to", max(years)))
        # add the axis labels
        title(
            xlab = "Year",
            ylab = "Life Expectancy (years)",
            col.lab = "#ececec",
            mgp = c(2.5, 0.5, 0)
        )
        # add mean line
        abline(h = mean_life_expectancy, col = "#ffffff67", lwd = 2, lty = 2)
        # add mean life expectancy line label
        text(
            x = max(years) - 2,
            y = mean_life_expectancy + 0.5,
            labels = paste(
                "Mean:",
                round(mean_life_expectancy, 2)),
            cex = 0.7,
            col = "#ececec"
        )
        box(col = "#767676")
        axis(1, col = "#767676", col.axis = "#ececec")
        axis(2, col = "#767676", col.axis = "#ececec")
    }
}
```

### Bonus: Similar, but using `tidyverse` and `ggplot2` (with default plotting options)

```{r task3_bonus}
# find countries starting with M, and mean life expectancy under 50
m_countries <- gapminder %>%
    filter(grepl("^M", country)) %>%
    group_by(country) %>%
    summarise(
        mean_life_expectancy = mean(lifeExp, na.rm = TRUE),
        years = paste0(min(year), "-", max(year)),
        year_min = min(year),
        year_max = max(year)
    ) %>%
    filter(mean_life_expectancy < 50) %>%
    select(country, years, mean_life_expectancy, year_min, year_max)

# print the countries found and their mean life expectancy
print(m_countries, n = Inf)


# plot the life expectancy (using facet wrap)
gapminder %>%
    filter(country %in% m_countries$country) %>%
    ggplot(aes(x = year, y = lifeExp, color = country)) +
    geom_line() +
    # add mean life expectancy line
    geom_hline(
        data = m_countries,
        aes(yintercept = mean_life_expectancy),
        linetype = "dashed",
        color = "white",
        alpha = 0.5
    ) +
    # add mean life expectancy line label
    geom_text(
        data = m_countries,
        aes(
            x = year_max,
            y = mean_life_expectancy + 1,
            label = paste("Mean:", round(mean_life_expectancy, 2))
        ),
        nudge_x = -2.5,
        color = "white",
        size = 2.25
    ) +
    theme_hc(style = "darkunica") +
    theme(strip.background = element_rect(colour = "#00000000",
                                        fill = "#00000065"),
          strip.text = element_text(color = "#ececec")) +
    facet_wrap(~ country, ncol = 2) +
    labs(
        title = "Life Expectancy for Countries Starting with M",
        subtitle = "Mean Life Expectancy under 50 years",
        x = "Year",
        y = "Life Expectancy (years)"
    )
```


