---
title: "Scraping police-killing data with R"
author: "Luke Ring"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    css: htmldoc.css
    highlight: breezedark
    theme: darkly
    toc: yes
    toc_depth: 2
    fig_height: 7
    fig_width: 10
    fig_retina: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Goal

> Use the rvest library to scrape data of your interest (football statistics in Wikipedia?, gender representatives in different governments? global population by country in https://www.worldometers.info/world-population/population-by-country/ )

I will look at the population changes since 1950 for the top 10 most populous countries in the world.

# Solution
First, install a handful of classic R packages and load their libraries:

- `rvest` for web-scraping
- `tidyverse` for data-wrangling, transformation and plots
- `stringr` for string manipulation
- `janitor` for clean headers that your OCD will love you for


```{r libraries, warning=FALSE, message=FALSE}
library(rvest)
library(tidyverse)
library(stringr)
library(janitor)
library(ggthemes)
```


## Scrape the data

First we need to find the top 10 most populous countries.

```{r url}
url <- "https://www.worldometers.info/world-population/population-by-country/"
# scrape the website
parsed_html <- read_html(url)
```

```{r scrape-table}
population_table <- parsed_html %>%
  html_elements("table") %>%
  html_table()

# get the actual data from the table
population_table <- population_table[[1]]

# peek at the data
head(population_table)

# now limit to the top 10 countries
population_table <- population_table[1:10, ]

# give the headers nice names
population_table <- population_table %>%
  clean_names() %>%
  rename(country = country_or_dependency, population = population_2020)

# rename "number" column to "year" and set to 2020
population_table <- population_table %>%
  rename(year = number) %>%
  mutate(year = 2020)

# peek at the cleaned
head(population_table)
```

Now we need to get the data for each of the top 10 countries.

```{r scrape-data}

add_country_col <- function(x) {
  # extract data
  x <- x %>%
    magrittr::extract2(1) %>%
    clean_names()
  # now get current country name
  country <- colnames(x)[13] %>%
    str_remove("_global_rank")
  # add country column
  x <- x %>%
    select(1:11) %>%
    mutate(country = country)
}

# get the country names
country_names <- population_table$country

# for the links United States is US
country_names <- str_replace(country_names, "United States", "US")

# get the country links
country_links <- country_names %>%
  str_replace_all(" ", "-") %>%
  str_replace_all("\\.", "") %>%
  str_to_lower() %>%
  paste0("https://www.worldometers.info/world-population/", ., "-population/")

population_data_css_selector <- "table.table-list:first-of-type"

# get the population data for each country, add the country name as a column and
# combine into a single data frame, matching columns
population_data <- country_links %>%
  map(read_html) %>%
  map(html_nodes, population_data_css_selector) %>%
  map(html_table, convert = FALSE) %>%
  map_df(add_country_col)

# peek at the data
head(population_data)

replace_pct <- function(x) {
  x <- str_replace_all(x, "%", "")
  x <- str_replace_all(x, " ", "")
  x <- as.numeric(x)
}

replace_thousands <- function(x) {
  x <- str_replace_all(x, ",", "")
  x <- as.numeric(x)
}

# now we can convert the data to numeric
population_data <- population_data %>%
  mutate_at(c("yearly_percent_change", "urban_pop_percent", "countrys_share_of_world_pop"), replace_pct)

population_data <- population_data %>%
  mutate_at(c("population", "yearly_change", "migrants_net", "median_age", "fertility_rate", "density_p_km2", "urban_population"), replace_thousands)

population_data$year <- as.numeric(population_data$year)

# replace u_s to united states
population_data$country <- str_replace(population_data$country, "u_s", "united states")
# make country names start with a capital letter
population_data$country <- str_to_title(population_data$country)

# peek at the data
head(population_data)
```

### Write result to file

Now we can write the data to a csv file.

```{r write-to-csv}
write_csv(population_data, "data/top10_country_historical_population_data.csv")
```

## Plot the data

Now we can plot the data.

```{r plot-data}
# define colors
country_colors <- c("#FF0000", "#FFA500", "#FFFF00", "#008000", "#0000FF", "#4B0082", "#EE82EE", "#6ccce9", "#808080", "#FFFFFF")

# overall population
population_data %>%
  ggplot(aes(x = year, y = population, color = country)) +
  geom_line() +
  scale_color_manual(values = country_colors) +
  labs(
    title = "Population of the top 10 most populous countries in the world",
    subtitle = "Data from https://www.worldometers.info/world-population/population-by-country/",
    x = "Year",
    y = "Population (millions)"
  ) +
  theme_hc(style = "darkunica") +
  theme(legend.position = "bottom")

# population growth
population_data %>%
  ggplot(aes(x = year, y = yearly_percent_change, color = country)) +
  geom_line() +
  scale_color_manual(values = country_colors) +
  labs(
    title = "Population growth of the top 10 most populous countries in the world",
    subtitle = "Data from https://www.worldometers.info/world-population/population-by-country/",
    x = "Year",
    y = "Population growth (percent change)"
  ) +
  theme_hc(style = "darkunica") +
  theme(legend.position = "bottom")

# by share of world population
population_data %>%
  ggplot(aes(x = year, y = countrys_share_of_world_pop, color = country)) +
  geom_line() +
  scale_color_manual(values = country_colors) +
  labs(
    title = "Share of world population of the top 10 most populous countries in the world",
    subtitle = "Data from https://www.worldometers.info/world-population/population-by-country/",
    x = "Year",
    y = "Share of world population (percent)"
  ) +
  theme_hc(style = "darkunica") +
  theme(legend.position = "bottom")

# by median age
population_data %>%
  ggplot(aes(x = year, y = median_age, color = country)) +
  geom_line() +
  scale_color_manual(values = country_colors) +
  labs(
    title = "Median age of the top 10 most populous countries in the world",
    subtitle = "Data from https://www.worldometers.info/world-population/population-by-country/",
    x = "Year",
    y = "Median age (years)"
  ) +
  theme_hc(style = "darkunica") +
  theme(legend.position = "bottom")

# by fertility rate
population_data %>%
  ggplot(aes(x = year, y = fertility_rate, color = country)) +
  geom_line() +
  scale_color_manual(values = country_colors) +
  labs(
    title = "Fertility rate of the top 10 most populous countries in the world",
    subtitle = "Data from https://www.worldometers.info/world-population/population-by-country/",
    x = "Year",
    y = "Fertility rate"
  ) +
  theme_hc(style = "darkunica") +
  theme(legend.position = "bottom")

# by density
population_data %>%
  ggplot(aes(x = year, y = density_p_km2, color = country)) +
  geom_line() +
  scale_color_manual(values = country_colors) +
  labs(
    title = "Density of the top 10 most populous countries in the world",
    subtitle = "Data from https://www.worldometers.info/world-population/population-by-country/",
    x = "Year",
    y = "Density (people per km2)"
  ) +
  theme_hc(style = "darkunica") +
  theme(legend.position = "bottom")

# by urban population
population_data %>%
  ggplot(aes(x = year, y = urban_pop_percent, color = country)) +
  geom_line() +
  scale_color_manual(values = country_colors) +
  labs(
    title = "Urban population of the top 10 most populous countries in the world",
    subtitle = "Data from https://www.worldometers.info/world-population/population-by-country/",
    x = "Year",
    y = "Urban population (percent)"
  ) +
  theme_hc(style = "darkunica") +
  theme(legend.position = "bottom")
```
